---
title: 'Title'
subtitle: 'Subtitle'
author: "Author"
date: '5 August 2025'
output:
  pdf_document:
    fig_width: 6
    fig_height: 3.75
    fig_caption: true
    highlight: zenburn
  html_document:
    fig_width: 6
    fig_height: 3.75
    fig_caption: true
    highlight: zenburn
  toc: yes
  toc_depth: 2
---

<!-- 
This is a comment in the .Rmd-file. 
It won't be displayed when you publish ("Knit") the file to html or pdf.
-->


<!-- ----------------------------------------------------------------------- -->
<!-- SOME SETTINGS & NOTES -->
<!-- ----------------------------------------------------------------------- -->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.pos = "!H", out.extra = "")
#knitr::opts_chunk$set(fig.width=5, fig.height=3)
```

```{=html}
(this is a longer comment in the .Rmd-file)

Markdown notes:  

 https://rmarkdown.rstudio.com/authoring_basics.html
 https://medium.com/@delucmat/specify-latex-parameters-in-r-markdown-9804467caaf 

- can comment out text by highlighting and pressing cmd+shift+C

- cannot use itemize or enumerate and have well-indented R-chunks in-between. Then, need to use + or - or 1. (2., etc.) in front of sentences that need to be itemized, and indent R chunks properly (one more than the section where it should stand). This seems to be the cleanest way of dealing with lists + r-chunks. Note: need to have an empty line before and after 1., 2.
Alternative: just use latex-enumerate/itemize and do not care about indentation.

- set figure settings for single chunk as : {r ..., fig.height = 15*0.75, fig.width = 16*0.75} (important e.g. for multiple plots in one (e.g. patchwork package))

```

<!-- 
Below we define some latex-settings (a new command in this case).
-->

\newcommand{\mean}{\mathbb{E}}

<!-- 
If you have a lot of such settings, it is best to write them in a separate file 
and store it as a .sty-file (latex-package), e.g. "mysettings.sty".
After that, you can add the following to your preamble (the very beginning of this document), 
say right after the line "date: ..." :

header-includes:
   - \usepackage{mysettings}
   
-->



<!-- ----------------------------------------------------------------------- -->
<!-- HERE THE DOCUMENT STARTS -->
<!-- ----------------------------------------------------------------------- -->


```{r libraries, echo = FALSE, warning=FALSE, message=FALSE}

rm(list = ls())  # delete everything in workspace/environment
cat("\014")      # delete everything in console


# Specify some path:

sDataPath <- "~//Dropbox/Coding/CodingSoftware/R/Templates"
setwd(sDataPath)


# Load some packages:

library(readxl)


```


# Section

This is an .Rmd-file. It integrates Markdown-text (incl. equations) and R-code.

## Subsection

Some text. Some inline-equation: $\theta=2$.

Another beautiful equation:
$$ x_i = \theta + u_i \; , \quad \text{with} \quad \mean[u_i|\theta]=0 \; .$$

## Another Subsection

**Some text in bold**

*Some text in italics*

```{r, echo=TRUE, results=FALSE, warning=FALSE, message=FALSE, tidy=TRUE, class.source="code"}

# A comment and some code:

a = 4
b = 3

a*b

```


1. First Item

```{r, echo=TRUE, results=FALSE, warning=FALSE, message=FALSE, tidy=TRUE, class.source="code"}

# A bit of code

c = 5

a*c

```


2. Second Item 

```{r, echo=TRUE, results=FALSE, warning=FALSE, message=FALSE, tidy=TRUE, class.source="code"}

# A bit of code

d = 8

a*d

```


## R-Markdown (.Rmd) vs Knitr (.Rnw)

Whereas .Rmd-files integrate Markdown-text and R-Code, .Rnw-files integrate latex-text and R-Code.
I prefer .Rmd-files. 
With them, you can generate both .html- and .pdf-files, whereas .Rnw only gives you .pdf-files.
Also, I find the error messages of .Rmd-files easier to read. 
In my experience, the only disadvantage of .Rmd-files relative to .Rnw-files is that they can't generate automatic enumerate-environments with R-code in between, but one has to write manually 1. , 2. , etc.

More generally, I think that the integration of text and R-code in a single document is useful for reports where the focus is on the coding (e.g. problem set solutions, data exploration, etc.), for which Markdown is good enough. 
For reports where the focus is on substance, there is typically no need to show your R-code integrated with latex-code in the document itself. Therefore, instead of using an .Rnw-file, you can set up a proper .tex-file and just include the output of your R-code (figures, tables, etc.) in it.

