---
title: 'R: Tutorial & Command-Collection'
author: | 
  | Source: https://github.com/markomlikota/CodingSoftware 
  | MIT License, © Marko Mlikota, https://markomlikota.github.io
  
output:
  html_document: 
    fig_width: 6
    fig_height: 3.75
    fig_caption: true
    theme: "simple"
    highlight: "zenburn"
    toc: false
    toc_depth: 1
    toc_float: true
    number_sections: false
  pdf_document:
    fig_width: 6
    fig_height: 3.75
    fig_caption: true
    theme: "simple"
    highlight: "zenburn"
    toc: false
    toc_depth: 1
    toc_float: true
    number_sections: false
---


<style type="text/css">
body, td {
   font-size: 18px;
}
code.r{
  font-size: 16px;
}
pre {
  font-size: 16px
}
h1 {
  font-size: 35px;
  margin: 60px 0px 30px 0px;
}
h2 {
  font-size: 25px;
}
#TOC {
  color: black; 
}

</style>

<script>
  var height=$('#primary').height(); // Calculate primary wrapper height
  $('#secondary').height(height); // Set the height it to sidebar
</script>




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(results=TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(tidy=TRUE)
knitr::opts_chunk$set(fig.pos = "!H", out.extra = "")
#knitr::opts_chunk$set(fig.width=5, fig.height=3)
```


# Plotting

```{r, echo=FALSE}

rm(list = ls()) # delete everything in workspace/environment
cat("\014")	# delete everything in console     


```

```{r, results=FALSE}

# A primary reason behind the popularity of R is the fact that one can create plots very flexibly. 
# Before we discuss more advanced plotting-techniques based on the package "ggplot2", 
# let's start with some basic plots that can be constructed using pre-built commands in R. 
# They are useful for checking quickly how something looks in the data-exploratory process 
# (and for inclusion in preliminary reports).

# Generate data to be plotted:

set.seed(50)

vx = seq(0.1,7,by=0.1)
vNormPDF = dnorm(vx,4,2)
vChi2PDF = dchisq(vx,3)
vtPDF = dt(vx,10)

library(mvtnorm)
vMeans = c(2,5,7)
mVariance = diag(c(2,1,3))
mVariance[1,2] = 1
mVariance[2,1] = 1
mNormDraws = rmvnorm(30,vMeans,mVariance)

# Load further data from actual dataset:
library(carData)
data(Hartnagel)
data(mtcars)

```


## Simple Plots Based on Built-In-Functions

```{r, eval=FALSE, fig.show='hide'}

# Plot single line:

# default:
plot(vNormPDF)

# to get line instead of dots:
plot(vNormPDF, type='l')
# for line and dots:
plot(vNormPDF, type='o')


# Note that by default, index is on the x-axis (1,2,...).
# To specify x-axis:
plot(vx,vNormPDF, type='l')


# specify y-limits, and x-limits: 
plot(vx, vNormPDF, type='l', ylim=c(0,0.25), xlim=c(-2,9)) 
# specify title, x-axis-label and y-axis-label:
plot(vx, vNormPDF, type='l', main="My Title Here", xlab="my x lab", ylab="my y lab") 
# note: can also be used to delete default labs: (default has no title)
plot(vx, vNormPDF, type='l', xlab="", ylab="") 


# specify color of markers (line and/or dots):
# (most standard colors are supported)
plot(vx, vNormPDF, type='l', col='blue') 
plot(vx, vNormPDF, col='blue')

# Make line thicker:
plot(vx, vNormPDF, type='l', lwd=4)


# Remove ticks for y-axis: 
plot(vx, vNormPDF, type='l', yaxt='n') 
# Remove ticks for x-axis:
plot(vx, vNormPDF, type='l', xaxt='n') 
# (these commands make more sense if the corresponding axis-label is also deleted)

```

```{r, eval=FALSE, fig.show='hide'}

# With package "latex2exp", we can add math to labels (x-lab, y-lab, title):

library(latex2exp)
sXlab = TeX("$\\theta_t $ (I made it up)")
sTitle = TeX("$f(\\theta_t)$")
plot(vx, vNormPDF, type='l', xlab=sXlab,ylab="",main=sTitle)

# The latex-part of the expression starts and ends with a dollar sign.
# The code is just as in latex, except that one has to use two backlashes to call latex-symbols.

```

```{r}

# Taken together, nice single-line plot:

plot(vx, vNormPDF, type='l', col='blue', lwd=4,
     main="my pretty plot of N(4,4)", xlab="", ylab="", yaxt='n',
     ylim=c(0,0.22), xlim=c(-1,8)) 

```

```{r, eval=FALSE, fig.show='hide'}

# We can easily add further lines to such an existing plot:

# Initial plot:
plot(vx, vNormPDF, type='l', col='blue')

# Add another line (chi^2-disttribution, in red):
lines(vx, vChi2PDF, type='l', col='red') 

# Note that the y-axis is not adjusted automatically so that the new line fits.


# Add mathematically defined lines:
abline(v=3,col='green')  # add vertical line at x=3
abline(h=0.08,col='gray') # add horizontal line at y=0.08
abline(a = -0.2, b = 0.1, col = "orange") # add line with specific intercept and slope

```

```{r}

# Taken together, nice lines plot:

plot(vx, vNormPDF, type='l', col='blue', lwd=4,
     main="my pretty plot of N(4,4) and Chi2(3)", xlab="", ylab="",
     ylim=c(0,0.3), xlim=c(-1,8)) 
lines(vx, vChi2PDF, type='l', col='red', lwd=4) 

```

```{r, eval=FALSE, fig.show='hide'}

# Note that we can also directly plot a function, 
# without first manually specifying a grid of x-values and evaluating the function at those values:

# Plot standard normal pdf:
plot(dnorm, xlim=c(-2,8))

# Plot pdf of N(4,4):
plot(function(x){ dnorm(x,4,2)}, xlim=c(-2,8))

# Plot exponential function:
plot(exp, xlim=c(1,3)) 

# Of course, all of the options discussed above can also be applied for these plots.
# They also work for all of the plots discussed below (scatter, histogram, boxplot, ...).
```


```{r, eval=FALSE}

# Saving a plot as a .png or .pdf file:

# 1. specify your path:
sMyPlotPath = "/Users/markomlikota/Dropbox/myfolder"
setwd(sMyPlotPath)

# 2. specify name:
png("blabla.png") # specify width and height here for nice plots

# 3. create the plot (or print an existing, already saved one with "print()"):
plot(vx, vNormPDF, type='l', col='blue', lwd=4,
     main="my pretty plot of N(4,4) and Chi2(3)", xlab="", ylab="",
     ylim=c(0,0.3), xlim=c(-1,8)) 
lines(vx, vChi2PDF, type='l', col='red', lwd=4) 

# 4. tell R that you're done: (no further lines or so will be added)
dev.off()


# Note that it is not possible, without rather complicated adjustments, to save such a plot as an object.
# This is different for plots created via "ggplot2".

```

```{r, eval=FALSE, fig.show='hide'}

# Scatter-Plot:
	
# To obtain a scatter plot, simply specify two vectors (as before) 
# and use dots rather than a line as markers:
plot(mNormDraws[,1],mNormDraws[,2])

# Change marker type:
plot(mNormDraws[,1],mNormDraws[,2], pch=8)
plot(mNormDraws[,1],mNormDraws[,2], pch=15)
# (see e.g. https://www.math.ucla.edu/~anderson/rw1001/library/base/html/points.html
# for further marker-types)


# add further cloud of points in different color:
points(mNormDraws[,1],mNormDraws[,3], col='red')

# add regression line for first cloud of points:
abline(lm(mNormDraws[,2] ~ mNormDraws[,1]))
# for second:
abline(lm(mNormDraws[,3] ~ mNormDraws[,1]), col = 'red')

# Should enlarge x- and y-axis to make this more appealing.

```

```{r}

# Taken together, nice scaatter-plot:

plot(mNormDraws[,1],mNormDraws[,2], col='blue', pch=4, lwd=2,
     main="my pretty scatterplot of Normal draws", xlab="", ylab="",
     ylim=c(2,11), xlim=c(-2,5)) 
points(mNormDraws[,1],mNormDraws[,3], col='red', pch=4, lwd=2)
abline(lm(mNormDraws[,2] ~ mNormDraws[,1]), col = 'blue')
abline(lm(mNormDraws[,3] ~ mNormDraws[,1]), col = 'red')

```

```{r, eval=FALSE, fig.show='hide'}

# Histogram:

# Default:
hist(mNormDraws[,1])

# Add color:
hist(mNormDraws[,1], col='blue')
# Add border color:
hist(mNormDraws[,1], col='blue', border='white')
# Remove fill-color:
hist(mNormDraws[,1], col=NULL, border='blue')


# Use density rather than frequency on y-axis: (so that area of bins sums up to 1)
hist(mNormDraws[,1],freq=FALSE)

# Specify number of bins by specifying location of bin-breaks:
hist(mNormDraws[,1],breaks=seq(-2,6,by=0.5))
# Specify number of bins: 
hist(mNormDraws[,1],breaks=8)
# (doesn't work perfectly; R uses it only as suggestion; see ?hist)

```

```{r}

# Taken together, nice histogram:

hist(mNormDraws[,1],breaks=seq(-1,5,by=0.5), col='blue', border='white',
     main="my pretty histogram of Normal draws", xlab="", ylab="")

```

```{r, eval=FALSE, fig.show='hide'}

# Boxplot:

# Boxplot of first vector of Normal draws:
boxplot(mNormDraws[,1]) 
# of first two:
boxplot(mNormDraws[,1],mNormDraws[,2]) 
# of all three: (i.e. of all columns in supplied matrix)
boxplot(mNormDraws) 

# Naming columns in that matrix changes x-tick-labels:
colnames(mNormDraws) <- c("first","second","third")
boxplot(mNormDraws) 


# With dataframes, one can directly choose to get 
# boxplots of one variable as a function of (values of) another:
boxplot(hp ~ cyl, data=mtcars, xlab="cyls",ylab="",main="Boxplot of HP by #cyls") 
# (rather than first manually creating vectors that show hp for each value of cyl)

```

```{r, eval=FALSE, fig.show='hide'}

# QQ-Plot:

qqnorm(vtPDF, xlab = "N(0,1)", ylab = "t(10)", col = "blue")
qqline(vtPDF, col="red")

```

```{r, eval=FALSE, fig.show='hide'}

# Display multiple plots as panels in a single figure:

library(dyn)

par(mfrow = c(1, 3)) # 1 x 3 plot
hist(mNormDraws[,1],breaks=seq(-1,8,by=1), col='blue', border='white',
     main="hist var1", xlab="", ylab="")
hist(mNormDraws[,2],breaks=seq(-1,8,by=1), col='blue', border='white',
     main="hist var2", xlab="", ylab="")
plot(mNormDraws[,1],mNormDraws[,2], col='blue', pch=4, lwd=2,
     main="scatter", xlab="var1", ylab="var2",
     ylim=c(2,11), xlim=c(-2,5)) 
abline(lm(mNormDraws[,2] ~ mNormDraws[,1]), col = 'blue')
par(mfrow = c(1, 1))


# Adjust margins and specify overarching title:

par(mfrow = c(1, 3), oma = c(2,2,2,2),mar = c(2,2,2,2))
# (see https://www.r-graph-gallery.com/74-margin-and-oma-cheatsheet.html)
hist(mNormDraws[,1],breaks=seq(-1,8,by=1), col='blue', border='white',
     main="hist var1", xlab="", ylab="")
hist(mNormDraws[,2],breaks=seq(-1,8,by=1), col='blue', border='white',
     main="hist var2", xlab="", ylab="")
plot(mNormDraws[,1],mNormDraws[,2], col='blue', pch=4, lwd=2,
     main="scatter", xlab="var1", ylab="var2",
     ylim=c(2,11), xlim=c(-2,5)) 
abline(lm(mNormDraws[,2] ~ mNormDraws[,1]), col = 'blue')
mtext("Bivariate Normal Draws", outer = TRUE, cex = 1.5)
par(mfrow = c(1, 1))

```


```{r, eval=FALSE, fig.show='hide'}

# Empty plot: (useful when creating e.g. a plot with 3 x 3 subplots and you have only 8 plots)
plot(c(0,0),type='n',ylab="",xlab="",axes=FALSE)


```

```{r, eval=FALSE, fig.show='hide'}


# Plotting xts objects:

# When plotting xts objects, we get a plot with lots of default-adjustments 
# (R realizes that the x-axis refers to time, returns a grid, etc.):

# To illustrate this, we download data on GDP and inflation directly from the FRED database:
library(quantmod)
# Inflation (based on GDP deflator):
getSymbols('GDPDEF',src='FRED')
GDPDEF = GDPDEF["1960-01-01/2025-01-01"] 
vInfl = 100 * 4 * diff(log(GDPDEF))[-1]
# GDP growth:
getSymbols('GDP',src='FRED')
GDP = GDP["1960-01-01/2025-01-01"] 
vGDPg = 100 * 4 * diff(log(GDP))[-1]


plot(vInfl, main = "Inflation (quarterly, annualized, %)")

# delete y-axis ticks on the right:
plot(vInfl, main = "Inflation (quarterly, annualized, %)", yaxis.right = FALSE)


# Plot two series on same graph:

plot(cbind(vInfl,vGDPg), main = "Inflation & GDP Growth")

# (better than manually;)
# plot(vInfl, main = "Inflation & GDP Growth", yaxis.right = FALSE)
# lines(vGDPg, col='red', yaxis.right = FALSE)

```


## Plots Based on "ggplot2"

```{r, results=FALSE, fig.show='hide'}


# With the package "ggplot2", one has much more flexibility in creating plots.

# We'll start with a plot containing a single line and then discuss 
# multiple lines, scatter plots, histograms and other plot-types. 
# Various plot-options are discussed, and each is introduced as part of the discussion of 
# the plot-type for which it is best explained.

# (A further popular package for plotting in R is "plotly". 
# However, in my view "ggplot2" is better in all aspects except for creating interactive plots.)

```


### Single-Line Plot

```{r, results=FALSE, fig.show='hide'}



library(ggplot2)

mData = data.frame(xvar=vx,norm=vNormPDF)

```

```{r, eval=FALSE, fig.show='hide'}

# Plot single line based on two vectors:

ggplot() + geom_line(aes(x=vx, y = vNormPDF))


# Plot single line based on a dataframe:

ggplot(data=mData) + geom_line(aes(x=xvar, y = norm))


# To change appearance of line, 
# specify further attributes in the option "geom_line()".
# e.g. change color, size and linetype:
ggplot(data=mData) + geom_line(aes(x=xvar, y = norm), color='blue', size=0.9, linetype="dashed")
# see http://sape.inf.usi.ch/quick-reference/ggplot2/linetype for linetypes available
# note: can specify linetype also as number (1 is solid, 2 is dashed, etc.)

# Add further options with pluses: 
ggplot(data=mData) + geom_line(aes(x=xvar, y = norm), color='blue', size=0.9) +
	labs(x="my x-lab", y="my y-lab", title= "my title") +
  theme_bw() + theme(aspect.ratio=5/8)

```

```{r, eval=FALSE, fig.show='hide'}

# Save plot to object:

pp1 = ggplot(data=mData) + geom_line(aes(x=xvar, y = norm), color='blue', size=0.9) +
  labs(x="my x-lab", y="my y-lab", title= "my title") 

# Then, can add further options to already existing plot:
pp1 + theme_bw() + theme(aspect.ratio=5/8)

# And can save to new object (or override existing one):
pp = pp1 + theme_bw() + theme(aspect.ratio=5/8)


# You may want to re-use some options over and over again. 
# Instead of typing them every time, store them as a list and call them at once all together:

plotOptions = list(
  theme_bw(), 
  theme(aspect.ratio=5/8) 
  )

# Show the plot "pp" with these options:
pp1 + plotOptions
# Add these options to plot "pp1":
pp1 = pp1 + plotOptions

```

```{r, eval=FALSE, fig.show='hide'}

# Save plot as .png or .pdf file:

# 1. create plot or print/call existing one:
pp
# 2. save:
ggsave("myprettyplot.png") 

# save with advanced options:
ggsave("myprettyplot.png",width=15,height=15*(8/5),units="cm") 
# note: width/height should match previously specified aspect ratio

# Note: for inclusion in latex-document, save plot best as pdf


```

```{r, results=FALSE, fig.show='hide'}

# As before, can have latex-code for labels: 

library(latex2exp)

sXlab = TeX("$\\theta_t $ (I made it up)")
sTitle = TeX("$ f(\\theta_t) $")

```

```{r, eval=FALSE, fig.show='hide'}

pp1 + labs(x=sXlab, y="", title=sTitle)


```

```{r, eval=FALSE, fig.show='hide'}

# There are many more options to customize the x-axis, y-axis and title:

# Adjust limits and ticks of x-axis and limits of y-axis: 
pp + 
  scale_x_continuous(breaks=seq(-1,8,3), limits=c(-1,8)) +
	scale_y_continuous(limits=c(0,0.25))

# Equivalent to the latter option:
pp + ylim(c(0,0.25))
# Also equivalent:
pp + expand_limits(y=c(0,0.25))
# When changing only upper limit, type:
pp + expand_limits(y=c(NA,0.25))
# (analogously for only lower limit)


# Manually specify labels of ticks on x-axis:
pp + scale_x_continuous(limits=c(-1,8), breaks=c(0,3,4,6), labels = paste(c(0,3,4,6),"a",sep=""))
# note: by specifying empty labels for certain breaks, 
# we can have ticks without labels:
vLabs = paste(c(0,3,4,6),"a",sep="")
vLabs[2] = ""
pp + scale_x_continuous(limits=c(-1,8), breaks=c(0,3,4,6), labels = vLabs)

# Change axis values from decimals to percent (e.g. 0.01 becomes 1%):
# (just for illustration; it doesn't make any sense here)
pp + scale_y_continuous(labels = scales::percent) 

# Have x-axis in units of pi:
vxDivPi = vx/pi
pp + scale_x_continuous(limits=c(-1,8), breaks=pi*(0:3), labels = paste(0:3,"pi"))
# (and analogously for any other transformation)

# Force all tick-labels to have same, specified number of decimals 
scaleFUN = function(x) sprintf("%.3f", x)
pp + scale_y_continuous(limits=c(0,0.25), breaks=seq(0,0.25,by=0.05), lab=scaleFUN)
# (without this option, we'd have two decimals in this case)
# (the option is also useful in some cases where R by default gives 
# different ticks with different numbers of decimals )


# Remove axis ticks and tick-labels:
pp + theme( axis.ticks.y=element_blank() , axis.text.y = element_blank() )

# To remove axis-label, can use +labels(ylab="") as above, or:
pp + theme( axis.title.y=element_blank() )

# Change font size for x-axis label:
pp + theme( axis.title.x = element_text(size=12) ) 

# Change font size of tick labels and space between axis and tick labels:
pp + theme( axis.text.x = element_text(size=12, vjust=-1) )

# Rotate x-axis labels:
pp + theme(axis.text.x = element_text(angle = 30, hjust = 0.5, vjust=0.8)) 


# Change font size and family of title: 
pp + theme( plot.title = element_text(family = "sans", size = 14) )
# Change also margins:
pp + theme( plot.title = element_text(family = "sans", size = 14, margin=margin(0,0,2,0) ) )
# (instead of margin, can also specify vjust and hjust)


```

```{r, results=FALSE, fig.show='hide'}

# If you have time on the x-axis, use the option "scale_x_date()" rather than "scale_x_manual()":

# Obtain quarterly GDP growth and inflation for illustration:

library(quantmod)

getSymbols('GDPDEF',src='FRED')
GDPDEF = GDPDEF["1960-01-01/2025-01-01"]
vInfl = 100 * 4 * diff(log(GDPDEF))[-1]

getSymbols('GDP',src='FRED')
GDP = GDP["1960-01-01/2025-01-01"]
vGDPg = 100 * 4 * diff(log(GDP))[-1]

mDataTS = data.frame(date=index(vGDPg), GDPgrowth=as.numeric(vGDPg), inflation=as.numeric(vInfl))
# sidenote: without "as.numeric()", the column names would be given by 
# the column names of the xts-objects "vGDPg" and "vInfl"
# and cannot be overwritten


# Plot GDP growth:
ppTS = ggplot(mDataTS) + geom_line(aes(x=date,y=GDPgrowth),size=0.9) +
          scale_x_date(date_labels = "%Y", 
                       date_breaks="2 years", 
                       limits=as.Date(c("2000-01-01", "2020-01-01")))


# Sometimes, the breaks don't appear as desired. e.g.
ppTS = ggplot(mDataTS) + geom_line(aes(x=date,y=GDPgrowth),size=0.9) +
          scale_x_date(date_labels = "%Y", 
                       date_breaks="5 years", 
                       limits=as.Date(c("2000-01-01", "2025-01-01")))
# In such cases, it's a good idea to set "breaks" rather than "date_breaks",
# for which it's useful to load package "lubridate":
library(lubridate)
ppTS = ggplot(mDataTS) + geom_line(aes(x=date,y=GDPgrowth),size=0.9) +
          scale_x_date(date_labels = "%Y", 
                       breaks = seq(ymd("2000/01/01"), ymd("2025/01/01"), by = "5 years"),
                       limits=as.Date(c("2000-01-01", "2025-01-01"))) 


# The comments from the section on string-dates above apply;
# if your dates are not in the format required by R ("YYYY-MM-DD"), 
# you must first change them to this format using the command "as.Date()". e.g.
as.Date("31.01.2020", "%d.%m.%Y") 


```
  
```{r, eval=FALSE, fig.show='hide'}


# Further useful options: 


# Remove grid:
pp + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())  

# Remove panel border (the box around the main plot area):
pp + theme(panel.border = element_blank())  

# Remove panel border but add x- and y-axis lines:
pp + theme(panel.border = element_blank()) +
     theme(axis.line = element_line(color = "black",size=0.2)) 


# Add margins of 1cm all around plot:

pp +  theme(plot.margin = unit(c(1,1,1,1), "cm"))


# Shading of specified area:
# (useful for confidence intervals)
pp + geom_ribbon(aes(ymin=vNormPDF*0.9, ymax=vNormPDF*1.1, x=vx), alpha=0.2, fill='blue')
# Note. can add several layers:
pp + 
  geom_ribbon(aes(ymin=vNormPDF*0.9, ymax=vNormPDF*1.1, x=vx), alpha=0.2, fill='blue') +
  geom_ribbon(aes(ymin=vNormPDF*0.8, ymax=vNormPDF*1.2, x=vx), alpha=0.2, fill='blue')


# Annotate text at given (x,y)-point:
pp + annotate("text",label="abc",x=2,y=0.07)
# (could specify hjust and vjust for further customization)

# note: with date on x-axis, need to put wrapper "as.Date()" around date:
ppTS + annotate("text",label="abc",x=as.Date("2003-01-31"),y=0.1)


# Flip axes:
pp + coord_flip()

```

```{r, eval=FALSE}

# Note that the options can be combined, 
# and they must be if they belong to the same option-command;
# e.g. don't write 
theme(panel.border = element_blank()) 
#and then 
theme( axis.title.x = element_text(size=12) )
# but write 
theme( panel.border = element_blank(), axis.title.x = element_text(size=12) )

# Note that this can create issues for the use of a list of pre-defined options.
# Sometimes when the same option is called several times, 
# the order how they are called can change the outcome.

```

```{r}

# Taken together, nice single-line plots:

plotOptions = list(
  theme_bw(), 
  theme(aspect.ratio=5/8,
          axis.title.x = element_text(size=12), 
          axis.title.y = element_text(size=12),
          axis.text.x = element_text(size=12), 
          axis.text.y = element_text(size=12),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          plot.title = element_text(family = "sans", size = 14)
        )
  )

ggplot(data=mData) + plotOptions +
  geom_line(aes(x=xvar, y = norm), color='blue', size=0.9) +
	  labs(x=sXlab, y="", title= sTitle) +
    scale_x_continuous(breaks=seq(-1,8,3), limits=c(-1,8)) +
	  scale_y_continuous(limits=c(0,0.25)) + 
      geom_ribbon(aes(ymin=vNormPDF*0.9, ymax=vNormPDF*1.1, x=vx), alpha=0.2, fill='blue') +
      geom_ribbon(aes(ymin=vNormPDF*0.8, ymax=vNormPDF*1.2, x=vx), alpha=0.2, fill='blue') 

ggplot(mDataTS) + plotOptions +
  geom_line(aes(x=date,y=GDPgrowth),color="blue", size=0.9) +
    labs(x="", y="", title="US GDP Growth (quarterly, annualized)") +
          scale_y_continuous(breaks=seq(-30,30,by=10)) +
          scale_x_date(date_labels = "%Y", 
                       breaks = seq(ymd("2000/01/01"), ymd("2025/01/01"), by = "5 years"),
                       limits=as.Date(c("2000-01-01", "2025-01-01"))) 

```


### Scatter-Plot

```{r, results=FALSE, fig.show='hide'}

mData = data.frame(norm1=mNormDraws[,1],norm2=mNormDraws[,2])

```

```{r, eval=FALSE, fig.show='hide'}


# A scatter plot is created analogously to a line plot. 
# We just use the option geom_point rather than geom_line.


# Based on two vectors:

ggplot() + geom_point(aes(x=mNormDraws[,1],y=mNormDraws[,2]))


# Based on data frame:

ggplot(data=mData) + geom_point(aes(x=norm1, y = norm2))


# Change marker appearance:

pp1 = ggplot(data=mData) + 
  geom_point(aes(x=norm1, y = norm2), size=1.5, shape=4, color="blue")
# see http://sape.inf.usi.ch/quick-reference/ggplot2/shape for shapes available

# Some markers (21 - 25) allow for filling:
pp2 = ggplot(data=mData) + 
  geom_point(aes(x=norm1, y = norm2), size=1.5, shape=21, color="blue", fill="blue")


# Can of course add all other options from above, e.g.

pp1 = pp1 + plotOptions + labs(x="my x-lab",y="my y-lab",title="my nice scatter")


# Add labels to scatter-points:

vMyLabels = paste("a",1:nrow(mData),sep="") # some arbitrary labels

pp1 + geom_text(aes(x=norm1, y = norm2, label=vMyLabels), size=2, color="purple", hjust=1, vjust=-0.5)

# add labels only to some points (e.g. only first 4):
vMyLabels[5:length(vMyLabels)] = "" # can also set to NA
pp1 + geom_text(aes(x=norm1, y = norm2, label=vMyLabels), size=2, color="purple", hjust=1, vjust=-0.5)



# Add lines to plot:

#line with specified intercept and slope (e.g. could add 45-degree line):
pp1 + geom_abline(intercept = 3, slope = 2, size=1, color = "black", linetype="dotted")
#vertical line:
pp1 + geom_vline(xintercept = 2, size=1, color = "black", linetype="dotted")
#regression line:
pp1 + geom_smooth(aes(x=norm1, y=norm2), method=lm, se=FALSE, size=1, color = "black", linetype="dotted")
#regression line with 90% confidence bands:
pp1 + geom_smooth(aes(x=norm1, y=norm2), method=lm, se=TRUE, level=0.9, size=1, color = "black", linetype="dotted")
# (note: single linear regression with homoskedasticity...)

```

```{r}

# Taken together, nice scatter plot:

ggplot(data=mData) +
  geom_point(aes(x=norm1, y = norm2), size=2, shape=21, color="blue", fill="blue") + 
  geom_smooth(aes(x=norm1, y=norm2), method=lm, se=FALSE, size=0.5, color = "darkblue", linetype="solid") + 
  labs(x="first Normal",y="second Normal",title="Scatter of Bivariate Normal Draws") + 
  plotOptions


```


### Plot With Multiple Lines

```{r, results=FALSE, fig.show='hide'}


# Rather than supplying several vectors, 
# it is easier to supply a dataframe with all the lines one wishes to plot:

mData = data.frame(xvar=vx, norm=vNormPDF,chi2=vChi2PDF, t=vtPDF)

# Note: the names of the different variables will be the line-labels. 
# They are a bit more difficult to change than the labels of the x- and y-axes.

```

```{r, eval=FALSE, fig.show='hide'}


# TWO LINES


ggplot(mData, aes(x=xvar)) + 
  geom_line(aes(y=norm)) +
  geom_line(aes(y=t))

# we should distinguish the different lines by color or shape or linetype; 
# e.g.
ggplot(mData, aes(x=xvar)) + 
  geom_line(aes(y=norm), color="blue") +
  geom_line(aes(y=t), color="red")

```

```{r, results=FALSE, fig.show='hide'}


# MANY (>=3) LINES


# In this case, it's best to first put the dataframe into long format,
# (as there could be many lines, 
# and adding each one separately like above would be cumbersome)

library(reshape2)
mDat = melt(mData, id.vars="xvar")

# the different variables are now distinguished in the column "variable", 
# and their values are shown in the column "value"

```

```{r, eval=FALSE, fig.show='hide'}

# Plot the three lines, distinguishing them by color:
ggplot(mDat, aes(x=xvar, y=value, group=variable)) +
	geom_line(aes(color=variable), size=0.9) # could specify linetype (for all lines)

# distinguish by linetype:
ggplot(mDat, aes(x=xvar, y=value, group=variable)) +
	geom_line(aes(linetype=variable), size=0.9) # could specify color (for all lines)

# distinguish by both:
ggplot(mDat, aes(x=xvar, y=value, group=variable)) +
	geom_line(aes(color=variable, linetype=variable))

# distinguish by line-color, point-color and point-shapes:
ggplot(mDat, aes(x=xvar, y=value, group=variable)) +
	geom_line(aes(color=variable),size=0.5) +
	geom_point(aes(color=variable, shape=variable),size=1)

# Of course, we can also just add the same markers for all lines:
ggplot(mDat, aes(x=xvar, y=value, group=variable)) +
	geom_line(aes(color=variable), size=0.9) +
  geom_point(aes(color=variable),shape=1)


# Set linetypes manually:

ppL = ggplot(mDat, aes(x=xvar, y=value, group=variable)) +
	geom_line(aes(linetype=variable), size=0.9) + plotOptions

ppL + scale_linetype_manual(values=c("solid", "dashed", "dotdash"))
# see http://sape.inf.usi.ch/quick-reference/ggplot2/linetype for linetypes available
# note: can specify linetype also as number (1 is solid, 2 is dashed, etc.)

# Note: provided vector may be longer than the number of lines you have.
# That's why you can create a long vector with your preferred linetypes 
# and then just call this vector, even for plots with differing number of lines.

# When setting linetypes manually, 
# can specify/change line labels:
ppL + scale_linetype_manual(values=c("solid", "dashed", "dotdash"), 
                            labels=c("mycurve1", "mycurve2", "mycurve3"))


# Set colors manually:

ppC = ggplot(mDat, aes(x=xvar, y=value, group=variable)) +
	geom_line(aes(color=variable), size=0.9) + plotOptions

ppC + scale_color_manual(values=c("blue","red","purple"))

# again can specify/change line labels:
ppC + scale_color_manual(values=c("blue","red","purple"),
                         labels=c("mycurve1", "mycurve2", "mycurve3"))

```

```{r, results=FALSE, fig.show='hide'}

# more choice of colors is given by package "colorspace":
library(colorspace)  

# take three colors (evenly spaced) from palette "Blues 2":
vMyColors = sequential_hcl(3, palette = "Blues 2") 

```

```{r, eval=FALSE, fig.show='hide'}

ppC + scale_color_manual(values=vMyColors)

```

```{r, results=FALSE, fig.show='hide'}

# to avoid too light first color, construct a vector with more colors,
# and then take the few ones you need from 
# the beginning and middle of the colorspace:
vMyColors = sequential_hcl(7, palette = "Blues 2")[c(1,3,5)]

```

```{r, eval=FALSE, fig.show='hide'}

ppC + scale_color_manual(values=vMyColors)


# Similarly, we can specify the colors of the markers as well as lines:

ppCM = ggplot(mDat, aes(x=xvar, y=value, group=variable)) +
	geom_line(aes(color=variable),size=0.5) +
	geom_point(aes(color=variable),size=1) +
  plotOptions

ppCM + scale_color_manual(values=vMyColors)


# Set point-shapes manually:

ppM = ggplot(mDat, aes(x=xvar, y=value, group=variable)) +
	geom_line(aes(color=variable),size=0.5) +
	geom_point(aes(color=variable, shape=variable),size=1) +
  plotOptions

ppM + scale_shape_manual(values=c(7,9,24))

ppM + scale_shape_manual(values=c(7,9,24),
                         labels=c("mycurve1", "mycurve2", "mycurve3"))
# see http://sape.inf.usi.ch/quick-reference/ggplot2/shape for shapes available


```

```{r, eval=FALSE, fig.show='hide'}

# Legend options:

# Take plot for illustration:

ppC = ppC + scale_color_manual(values=vMyColors)


# Remove legend title:

ppC + theme( legend.title=element_blank() )


# Remove legend altogether:

ppC + theme(legend.position = "none") 


# Change position and direction:
# (by default, legend is outside the plot (middle-right) and vertical)

#put to middle-left:
ppC + theme( legend.position="left" )

#put to bottom-right:
ppC + theme( legend.justification="bottom" )

#make horizontal and put to bottom of plot:
ppC + theme(legend.direction="horizontal", legend.position="bottom", legend.justification="left")

#put inside the plot: 
ppC + theme( legend.position=c(0.8, 0.75) )
#( top right is (1,1), top left is (0,1))


# In some cases, it's hard to put the legend inside the plot, 
# but without crossing the lines/markers. e.g.
ppC + theme( legend.position=c(0.5, 0.5))
# Unfortunately, one cannot make the legend background fully transparent.
# Make legend-text transparent at least:
ppC + theme( legend.position=c(0.5, 0.5), legend.background=element_rect(fill = alpha("white", 0.0)) ) 


```

```{r}

# Taken together, nice multiple-line plot:

ggplot(mDat, aes(x=xvar, y=value, group=variable)) +
	geom_line(aes(color=variable), size=0.9) + plotOptions + 
  scale_color_manual(values=vMyColors) +
  theme( legend.title=element_blank(), legend.position=c(0.8, 0.75) ) +
  labs(x="",y="",title="Three Important Distributions")

```


### Scatter-plot with multiple variables

```{r, results=FALSE, fig.show='hide'}

# Analogous to above, just with geom_point instead of geom_line option.


# Create dataframe with variables (vectors) to plot:
mData = as.data.frame(mNormDraws)
colnames(mData) = c("norm1","norm2","norm3")

# Melt to long format:
mDat = melt(mData,id.vars = "norm1") 
 
# Note: variable taken here as id.var will be on x-axis.
# In a scatter plot, this could in principle be any variable.
# (In contrast, under the line plot, it was clear what the x-axis is.)

```

```{r, eval=FALSE, fig.show='hide'}

# Plot, distinguishing variables by markers:

ggplot(mDat, aes(x=norm1, y=value, group=variable)) +
	geom_point(aes(shape=variable), size=2, color="blue")

# Plot, distinguishing variables by colors:

ggplot(mDat, aes(x=norm1, y=value, group=variable)) +
	geom_point(aes(color=variable), size=2)

# Plot, distinguishing variables by both:

ggplot(mDat, aes(x=norm1, y=value, group=variable)) +
	geom_point(aes(color=variable, shape=variable), size=2)


# Add options:

ppMS = ggplot(mDat, aes(x=norm1, y=value, group=variable)) +
	geom_point(aes(color=variable, shape=variable), size=2) +
  plotOptions


# Specify marker-shapes and colors manually:

ppMS + scale_color_manual(values=vMyColors) + scale_shape_manual(values=c(1,4))

# Also re-label variables:
ppMS + scale_color_manual(values=vMyColors, labels=c("1st Normal","2nd Normal")) + 
  scale_shape_manual(values=c(1,4), labels=c("1st Normal","2nd Normal"))
# (note: just doing it in one of the two options will produce two legends)

# Note: when distinguishing by color,
# by default, R will choose a round marker that is filled (as above). e.g.
ggplot(mDat, aes(x=norm1, y=value, group=variable)) +
	geom_point(aes(color=variable), size=2)
# When you change the markers manually, like here, then they will not be filled.
# To add filling, must let filling also change by variable. 
# e.g. this produces unfilled triangles:
ggplot(mDat, aes(x=norm1, y=value, group=variable)) +
	geom_point(aes(color=variable), size=2, shape=2)
# to get filled triangles:
ggplot(mDat, aes(x=norm1, y=value, group=variable)) +
	geom_point(aes(color=variable,fill=variable), size=2, shape=24)


```

```{r}

# Nice, final scatter-plot with multiple variables:

ggplot(mDat, aes(x=norm1, y=value, group=variable)) +
	geom_point(aes(color=variable), size=2) +
  scale_color_manual(values=c("blue","red"), labels=c("2nd Normal","3rd Normal")) +
  plotOptions +
  scale_y_continuous(limits=c(0,12),breaks=c(0,4,8,12)) +
  theme( legend.title=element_blank(), legend.position=c(0.8, 0.15) ) +
  labs(x="1st Normal",y="",title="Multivariate Normal Draws (k=3)")



```


### "Grouped" Scatter-Plot

```{r, eval=FALSE, fig.show='hide'}


# Above, we plotted three variables in a single scatter plot, 
# i.e. we plotted two variables against a third.

# In the same way, we could have plotted, say, 5 variables 
# by plotting 4 variables against a fifth one, yielding 4 "clouds" of points.

# Sometimes, we indeed want to plot two variables,
# but distinguish the points in the scatter by a third variable.
# The number of "clouds" of points is then determined 
# by how many values this third variable has.


# DISCRETE CASE


# Plot hp against hp, with different colors for different numbers of cylinders:

ggplot(mtcars, aes(x=mpg, y=hp, group=cyl)) +
	geom_point(aes(color=cyl))

```

```{r, results=FALSE, fig.show='hide'}

# Note that R thinks of cyl as a continuous variable, even though it takes only three values.
# To fix that, code it as type "character" rather than "numeric":

mtcars2 = mtcars # create new dataframe..
mtcars2$cyl = as.character(mtcars2$cyl) #.. in which cyl is of type character

```

```{r, eval=FALSE, fig.show='hide'}

ppMS = ggplot(mtcars2, aes(x=mpg, y=hp, group=cyl)) +
	geom_point(aes(color=cyl))

# We can change the colors as above; e.g.
ppMS + scale_color_manual(values=vMyColors, 
                          labels=c("4 cyl.","6 cyl.","8 cyl."))
  

# Also, could of course distinguish points for different numbers of cylinders 
# by markers rather than colors (or both).

# In this case, cyl really needs to be of type character,
# because, in contrast to colors, markers cannot vary continuously.



# CONTINUOUS CASE


# Plot hp against mpg, and distinguish points by color based on value of qsec:
ppMS = ggplot(mtcars, aes(x=mpg, y=hp, group=qsec)) +
	geom_point(aes(color=qsec))

# To specify colors, need to use scale_color_gradient rather than scale_color_manual:
ppMS + scale_color_gradient(low="white",high="blue")

# The analogous holds for the filling 
# (if marker-shapes are specified that can be filled):
ppMS = ggplot(mtcars, aes(x=mpg, y=hp, group=qsec)) +
	geom_point(aes(color=qsec),shape=21) +
  scale_color_gradient(low="white",high="blue") +
  scale_fill_gradient(low="white",high="blue")


```

```{r}

# Nice, final grouped scatter-plot:

ggplot(mtcars, aes(x=mpg, y=hp, group=qsec)) +
	geom_point(aes(color=qsec,fill=qsec),size=3,shape=21) + 
  scale_color_gradient(low="white",high="blue") +
  scale_fill_gradient(low="white",high="blue") +
  plotOptions +
  scale_y_continuous(limits=c(0,300),breaks=seq(0,300,by=100)) +
  theme( legend.position="none" ) +
  labs(title="A Nice Grouped-Scatter Plot")


```


### Histogram

```{r, eval=FALSE, fig.show='hide'}

# SINGLE HISTOGRAM


# Based on single vector:

ggplot() + geom_histogram(aes(x=mNormDraws[,2]))

# Based on column in dataframe:
ggplot(data=mData) + geom_histogram(aes(x=norm2))

# note that here we only need to specify the x-variable in aes()


# Change y-axis from frequency to density:

ggplot() + geom_histogram(aes(x=mNormDraws[,2],y=..density..))

ggplot(data=mData) + geom_histogram(aes(x=norm2,y=..density..))


# Specify number of bins directly:
ggplot(data=mData) + geom_histogram(aes(x=norm2,y=..density..), bins=10)

# Specify number of bins with binwidth:
ggplot(data=mData) + geom_histogram(aes(x=norm2,y=..density..), binwidth=0.5)


# Change color and filling:
ggplot(data=mData) + geom_histogram(aes(x=norm2,y=..density..), color="black", fill="grey", alpha=0.2)


# Add line:
# (here: add actual pdf of N(5,1))
vThisNormPDF = dnorm(vx,5,1)
ggplot() + 
  geom_histogram(aes(x=mNormDraws[,2],y=..density..))+
  geom_line(aes(x=vx,y=vThisNormPDF))

# Could also do this with dataframe, but then we need same number of draws
# as "elements" in the vector that represents the line (here: function evaluations)



# MULTIPLE HISTOGRAMS IN SINGLE PLOT


ggplot(data=mData) + 
  geom_histogram(aes(x=norm2,y=..density..),color="blue",fill="blue",alpha=0.2) + 
  geom_histogram(aes(x=norm3,y=..density..),color="red",fill="red",alpha=0.2) 


# A separate histogram of one variable for each value of some third variable:
# e.g. a separate histogram of hp for each value of cyl, each with different color:

ggplot(mtcars2, aes(x=hp, fill=cyl)) + 
  geom_histogram(aes(x=hp, y = ..density.., color=cyl), position = 'identity')

# Should make bars a bit transparent:
ggplot(mtcars2, aes(x=hp, fill=cyl)) + 
  geom_histogram(aes(x=hp, y = ..density.., color=cyl), position = 'identity', alpha=0.6)
  

```


```{r}

# Taken together, nice histogram plots:

vThisNormPDF = dnorm(vx,5,1)

ggplot() + plotOptions +
  geom_histogram(aes(x=mNormDraws[,2],y=..density..), color="blue",fill="blue",alpha=0.2)+
  geom_line(aes(x=vx,y=vThisNormPDF),color="purple") +
  labs(x="",y="",title="Histogram of N(5,1) draws") +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank())

ggplot(mtcars2, aes(x=hp, fill=cyl)) + plotOptions +
  geom_histogram(aes(x=hp, y = ..density.., color=cyl), position = 'identity', alpha=0.6) +
  theme( legend.title=element_blank(), legend.position=c(0.8, 0.75)) +
  labs(x="",y="",title="Histogram of hp by cyl.") +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank())


```


### Bar-Plot

```{r, results=FALSE, fig.show='hide'}

# Compute mean of Normal draws for each 
# column of mNormDraws (dimension of multiv. Normal): 

mDataBP = data.frame(norms=c(1,2,3),means=apply(mNormDraws,2,mean))

```

```{r, eval=FALSE, fig.show='hide'}

# Illustrate: 

ggplot(data=mDataBP, aes(x=norms, y=means)) +
  geom_bar(stat="identity")

# Adjust appearance of bars:
ggplot(data=mDataBP, aes(x=norms, y=means)) +
  geom_bar(stat="identity",color="blue",fill="blue",alpha=0.2) 

```

```{r, results=FALSE, fig.show='hide'}

# To change x-axis, it is best to first make R realize that 
# it is a discrete rather than continuous variable:
mDataBP$norms = as.character(mDataBP$norms)

```

```{r, eval=FALSE, fig.show='hide'}

ggplot(data=mDataBP, aes(x=norms, y=means)) +
  geom_bar(stat="identity",color="blue",fill="blue",alpha=0.2) + 
  scale_x_discrete(labels=c("1st","2nd","3rd"))

```


```{r}

# Taken together, nice bar-plot:

ggplot(data=mDataBP, aes(x=norms, y=means)) + plotOptions +
  geom_bar(stat="identity",color="blue",fill="blue",alpha=0.2) + 
  scale_x_discrete(labels=c("1st","2nd","3rd")) +
  labs(x="",y="",title="Mean of Normal Draws")
  
```


### Plot-Combinations Into Single Figure

```{r, eval=FALSE, fig.show='hide'}

# Note: when producing a latex document with these plots,
# it is a better idea to produce the plots individually and assemble them in latex.
# (more flexibility)

# Three plots for illustration:
plotOptionsCommonHere = 
  list(plotOptions,
       ylim(c(0,6)),
       scale_x_continuous(limits=c(-2,10),breaks=seq(-2,10,by=2))
       )
pp1 = ggplot(data=mData) + plotOptionsCommonHere +
        geom_histogram(aes(x=norm1),color="blue",fill="blue",alpha=0.2) + 
        labs(x="",y="",title="norm1") 
  
pp2 = ggplot(data=mData) + plotOptionsCommonHere +
        geom_histogram(aes(x=norm2),color="blue",fill="blue",alpha=0.2) + 
        labs(x="",y="",title="norm2") 

pp3 = ggplot(data=mData) + plotOptionsCommonHere +
        geom_histogram(aes(x=norm3),color="blue",fill="blue",alpha=0.2) + 
        labs(x="",y="",title="norm3") 


# It might be useful to create an empty plot:
# (e.g. when creating a 3x3 figure, but have 8 plots)

ggplot() + theme_void()



# FIRST APPROACH


library(gridExtra) 
library(ggpubr)
# (need both packages)

# Create 1 x 2 plot:
grid.arrange(pp1, pp2, ncol=2) 

# To save:
ggsave("myplot1.pdf", arrangeGrob( pp1, pp2, ncol=2) )


# Add title:
title = text_grob("Draws from two Normals", size = 14, face = "bold", vjust=5)
grid.arrange(pp1, pp2, ncol=2, top=title)

# Can specify widths:
grid.arrange(pp1, pp2, ncol=2, top=title, widths = c(2,2))

# Can also specify ordering of plots (useful when many more than just 2 plots are added):
grid.arrange(pp1, pp2, ncol=2, top=title, widths = c(2,2), layout_matrix = rbind(c(1,2))) 


# All of these options need to be specified also in "arrangeGrob()" when saving the plot.

# Note: it would make sense to delete y-axis ticks on the right plot 
# as they are common to both and already shown in the left plot.



# SECOND APPROACH


library(patchwork)

# Create 1 x 2 plot as above:
(pp1 | pp2)

# Here we have the advantage that we can save the resulting plot and call it later:
ppTotal = (pp1 | pp2)
ppTotal

# Add title:
ppTotal + plot_annotation(title = "Draws from two Normals")

# With this approach, we do not need a special command for saving the plots, 
# but can use "ggsave()".
# However, one needs to create a large plot,
# because with many plots, default size window is usually too small:
ggsave("myplot2.pdf",width=40,height=25,units="cm") 
# Note: changing the width and height will also change the spacing between figures.


# 2 x 3 plots:
(pp1 | pp1 | pp1) /
  (pp1 | pp1 | pp1)

# Can have even more flexible plot-arrangement:
layout = "
      AABB
      AABB
      EEEE
      EEEE
    "
pp1 + pp1 + pp1 + plot_layout(design = layout)
ggsave("myplot2.pdf",width=40,height=25,units="cm") 

# For plots with legends:
# If all plots have the same legend, 
# you can collect all them into one single legend using the following commands:
ppTotal + plot_layout(guides = "collect") + theme(legend.position = 'bottom')


```

### Further

```{r, eval=FALSE, fig.show='hide'}

# There are many more things that ggplot2 can do. 
# For example, to create a grid-plot, use "geom_raster()" or "geom_rect()", 
# to create an area-plot, use "geom_area()".
# A google search usually shows quick how to use these commands.


# A final, important note: 
# when iteratively adding layers to a ggplot in a loop 
# (e.g. adding one line at a time to an existing plot), 
# you need to use double quotation marks. 
# Otherwise, ggplot creates the plot only once the loop is finished, which may lead to unexpected behavior. 
# See this post:
# https://stackoverflow.com/questions/52671480/adding-multiple-ribbons-in-ggplot.

```



